---
title: Analysis of 2018 tornado data from NOAA
date: October 23, 2019
author: Peter Carbonetto
output:
  html_document:
    theme: readable 
---

This is my analysis of tornado data downloaded from the
[NOAA website][noaa-storm-events].

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",fig.height = 5,fig.width = 9)
```

# Load packages

Load the packages used for the analysis.

```{r load-pkgs, message=FALSE, warning=FALSE}
library(readr)
library(ggplot2)
library(cowplot)
```

# Load tornado data

Load the 2018 storm event data from the CSV file. The output should be
a data frame with 62,339 rows and 51 columns.

```{r load-data, message=FALSE, warning=FALSE}
storms <- read_csv("StormEvents_details-ftp_v1.0_d2018_c20190817.csv.gz")
class(storms) <- "data.frame"
nrow(storms)
ncol(storms)
```

Here we focus on events classified as tornadoes. There should be 1,250
weather events that are tornadoes.

```{r get-tornado-data}
tornadoes <- subset(storms,EVENT_TYPE == "Tornado")
nrow(tornadoes)
```

The tornadoes are categorized by a severity scale, with "EF3" being
the most severe. Only a small number (12) of the tornadoes 2018 were
classified as EF3.

```{r summarize-f-scale}
tornadoes <- transform(tornadoes,TOR_F_SCALE = factor(TOR_F_SCALE))
summary(tornadoes$TOR_F_SCALE)
```

# Plot geographic distribution of tornadoes

Using the geographic data provided (the latitudes and longitudes),
create a map showing the location of all tornado events that occurred
in 2018. Color the tornadoes by severity to highlight the most severe
tornadoes (shown in orange and magenta).

```{r plot-tornadoes}
rows      <- order(tornadoes$TOR_F_SCALE)
tornadoes <- tornadoes[rows,]
ggplot(tornadoes,aes(x = BEGIN_LON,y = BEGIN_LAT,color = TOR_F_SCALE)) +
  geom_path(data = map_data("state"),aes(x = long,y = lat,group = group),
            color = "black") +
  geom_point(shape = 20,size = 4) +
  scale_color_manual(values = c("lightskyblue","skyblue","darkorange",
                                "magenta","gray")) +
  labs(x = "longitude",y = "latitude")
```

As expected, the greatest concentration of severe tornadoes is in
Midwestern states such as Kansas and Mississippi. There is also a
quite large cluster of severe tornadoes in and around upper
Louisiana. Indeed, there was a ["tornado outbreak" that occurred in
Louisiana in April 2018][tornado-outbreak].

[noaa-storm-events]: https://www.ncdc.noaa.gov/stormevents/ftp.jsp
[tornado-outbreak]: https://en.wikipedia.org/wiki/Tornado_outbreak_of_April_13%E2%80%9315,_2018
